---
title: "Reproducible GSEA Benchmarking"
author:
    - name: Ludwig Geistlinger
      affiliation: School of Public Health, City University of New York
      email: ludwig.geistlinger@sph.cuny.edu
package: GSEABenchmarkeR
abstract: > 
    The _GSEABenchmarkeR_ package implements an extendable framework for
    reproducible evaluation of set- and network-based methods for enrichment
    analysis of gene expression data. This includes support for the efficient
    execution of these methods on comprehensive real data compendia (microarray
    and RNA-seq) using parallel computation on standard workstations and
    institutional computer grids. Methods can then be assessed with respect to
    runtime, statistical significance, and relevance of the results for the
    phenotypes investigated.
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  % \VignetteIndexEntry{Reproducible GSEA Benchmarking}
  % \VignetteEngine{knitr::rmarkdown}
---

```{r setup, echo=FALSE}
suppressPackageStartupMessages({ 
    library(GSEABenchmarkeR)
    library(EnrichmentBrowser)
})
```

# Purpose of the package

The purpose of the `r Biocpkg("GSEABenchmarkeR")` package is to compare the performance 
of different methods for gene set enrichment analysis across many gene expression 
datasets.
 
Users interested in conducting gene set enrichment analysis for a specific dataset
of choice are recommended to use the `r Biocpkg("EnrichmentBrowser")` package instead.

In other words,

- if you are interested in analyzing a particular microarray or RNA-seq dataset,
e.g. a case-control study where you want to find out which GO terms / KEGG pathways 
are enriched for differentially expressed genes, i.e. your primary goal is  
_biological interpretation of a specific dataset under investigation_, then use 
the `r Biocpkg("EnrichmentBrowser")` package.

- if you want to assess the performance (runtime, type I error rate, etc) of  
different enrichment methods across many datasets and in certain simulated setups - 
i.e. your primary goal is to  
_understand methodological aspects and compare methods against each other_,
then use the `r Biocpkg("GSEABenchmarkeR")` package.

# Setup
Although gene set enrichment analysis (GSEA) has become an integral part of 
high-throughput gene expression data analysis, the assessment of enrichment 
methods remains rudimentary and *ad hoc*.
In the absence of suitable gold standards, the evaluation is commonly restricted
to selected data sets and biological reasoning on the relevance of resulting 
enriched gene sets.
However, this is typically incomplete and biased towards a novel method being 
presented.

As the evaluation of GSEA methods is thus typically based on self-defined 
standards, [Mitrea et al. (2013)](https://doi.org/10.3389/fphys.2013.00278) 
identified the lack of gold standards for consistent assessment and comparison 
of enrichment methods as a major bottleneck.
Furthermore, it is often cumbersome to reproduce existing assessments for 
additional methods, as this typically involves considerable effort of data 
processing and method collection.

Leveraging the representative and extendable collection of
enrichment methods available in the `r Biocpkg("EnrichmentBrowser")` package, 
the `r Biocpkg("GSEABenchmarkeR")` package facilitates efficient execution 
of these methods on comprehensive real data compendia.
The compendia are curated collections of microarray and RNA-seq datasets 
investigating human diseases (mostly specific cancer types), for which 
disease-relevant gene sets have been defined _a priori_.

Consistently applied to these datasets, enrichment methods can then be subjected
 to a systematic and reproducible assessment of _(i)_ computational runtime, 
_(ii)_ statistical significance, especially how the fraction of 
significant gene sets relates to the fraction of differentially expressed genes,
and _(iii)_ phenotype relevance, i.e. whether enrichment methods produce gene set 
rankings in which phenotype-relevant gene sets accumulate at the top.
   
In the following, we demonstrate how the package can be used to

- load specific pre-defined and user-defined data compendia,
- carry out differential expression analysis across datasets,
- apply enrichment methods to multiple datasets, and
- benchmark results with respect to the chosen criteria.

We start by loading the package.

```{r lib}
library(GSEABenchmarkeR)
```

# Expression data sources

The `r Biocpkg("GSEABenchmarkeR")` package implements a general interface for 
loading compendia of expression datasets.
This includes 

- the pre-defined GEO2KEGG microarray compendium that consists of 42 datasets 
    investigating a total of 19 different human diseases as collected by 
    Tarca et al. ([2012](https://doi.org/10.1186/1471-2105-13-136) and
    [2013](https://doi.org/10.1371/journal.pone.0079217)), 
- the pre-defined TCGA RNA-seq compendium, consisting of datasets from 
    [The Cancer Genome Atlas](https://cancergenome.nih.gov) 
    investigating a total of 34 different cancer types, and 
- user-defined data from file.

In the following, we describe both pre-defined compendia in more detail and also
demonstrate how user-defined data can be incorporated.

## Microarray compendium

Although RNA-seq (read count data) has become the *de facto* standard for 
transcriptomic profiling, it is important to know that many methods for 
differential expression and gene set enrichment analysis have been originally
developed for microarray data (intensity measurements).
However, differences in data distribution assumptions (microarray: quasi-normal,
RNA-seq: negative binomial) have made adaptations in differential expression 
analysis and, to some extent also in gene set enrichment analysis, necessary.

Nevertheless, the comprehensive collection and curation of microarray data in 
online repositories such as [GEO](https://www.ncbi.nlm.nih.gov/geo) still 
represent a valuable resource.
In particular, Tarca et al. ([2012](https://doi.org/10.1186/1471-2105-13-136) 
and [2013](https://doi.org/10.1371/journal.pone.0079217)) compiled 42 datasets
from GEO, each investigating a human disease for which a specific 
[KEGG](http://www.genome.jp/kegg) pathway exists. 

These pathways are accordingly defined as the target pathways for the various 
enrichment methods when applied to the respective datasets. For instance, 
methods are expected to rank the 
[Alzheimer's disease pathway](http://www.genome.jp/kegg-bin/show_pathway?hsa05010)
close to the top when applied to
[GSE1297](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE1297),
a case-control study of Alzheimer's disease.    

Furthermore, Tarca et al. made these datasets available in the _Bioconductor_ 
packages `r Biocpkg("KEGGdzPathwaysGEO")` and 
`r Biocpkg("KEGGandMetacoreDzPathwaysGEO")`.

The `r Biocpkg("GSEABenchmarkeR")` package simplifies access to the compendium 
and allows to load it into the workspace via 

```{r maComp}
geo2kegg <- loadEData("geo2kegg")
names(geo2kegg)
```

A specific dataset of the compendium can be obtained via

```{r getDatasetProbe}
geo2kegg[[1]]
```

which returns, in this example, an `ExpressionSet` (documented in the 
`r Biocpkg("Biobase")` package) that contains expression levels of 
22,283 probe sets measured for 16 patients. 

To prepare the datasets for subsequent analysis, the `r Biocpkg("GSEABenchmarkeR")`
package provides the function `maPreproc`.
The function invokes `EnrichmentBrowser::probe2gene` on each dataset 
to summarize expression levels for probes annotated to the same gene. 
Here, we apply the function to the first 5 datasets of the compendium. 

```{r maPreproc}
geo2kegg <- maPreproc(geo2kegg[1:5])
```

Now, 
```{r getDatasetGene}
geo2kegg[[1]]
```
returns a `r Biocpkg("SummarizedExperiment")` that contains the summarized 
expression values for 12,994 genes. Furthermore, sample groups 
 are defined in the `colData` column **GROUP**, yielding here 7 cases (1) and 
9 controls (0).

```{r maGroups}
se <- geo2kegg[[1]]
table(se$GROUP)
```

Note: The `maPreproc` returns datasets consistently mapped to NCBI Entrez
Gene IDs, which is compatible with most downstream applications. However, 
mapping to a different ID type such as ENSEMBL IDs or HGNC symbols can also be 
done using the function `EnrichmentBrowser::idMap`.  

## RNA-seq compendium

The Cancer Genome Atlas ([TCGA](https://cancergenome.nih.gov)) project performed
 a molecular investigation of various cancer types on an unprecedented scale 
including various genomic high-throughput technologies.
In particular, transcriptomic profiling of the investigated cancer types has 
comprehensively been carried out with RNA-seq in tumor and adjacent normal tissue.

Among the various resources that redistribute TCGA data, 
[Rahman et al. (2015)](https://doi.org/10.1093/bioinformatics/btv377)
consistently preprocessed the RNA-seq data for 24 cancer types and made the data 
available in the GEO dataset
[GSE62944](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE62944).

The GSE62944 compendium can be loaded using the `loadEData` function, 
which provides the datasets ready for subsequent differential expression and 
gene set enrichment analysis.

Here, we load the compendium into the workspace using only two of the datasets. 

```{r rseqComp}
tcga <- loadEData("tcga", nr.datasets=2)
names(tcga)
```

For example, the breast cancer dataset contains RNA-seq read counts for roughly
20,000 genes measured in 1,119 tumor (1) and 113 adjacent normal (0) samples. 

```{r brca}
brca <- tcga[[2]]
brca
table(brca$GROUP)
``` 

## User-defined data compendium

With easy and fast access to the GEO2KEGG and TCGA compendia, enrichment 
methods can be directly applied and assessed on well-studied, standardized 
expression datasets.
Nevertheless, benchmarking with the `r Biocpkg("GSEABenchmarkeR")` package is 
designed to be extendable to additional datasets as well.
 
Therefore, the `loadEData` function also accepts a directory where datasets,
preferably of class `r Biocpkg("SummarizedExperiment")`, have been saved as
[RDS](https://stat.ethz.ch/R-manual/R-devel/library/base/html/readRDS.html)
files.

```{r userComp}
data.dir <- system.file("extdata", package="GSEABenchmarkeR")
edat.dir <- file.path(data.dir, "myEData")
edat <- loadEData(edat.dir)
names(edat)
edat[[1]]
```

# Differential expression

To perform differential expression (DE) analysis between sample groups for 
selected datasets of a compendium, the `r Biocpkg("GSEABenchmarkeR")` package 
provides the function `runDE`.

The function invokes `EnrichmentBrowser::deAna` on each dataset, which contrasts
the sample groups as defined in the **GROUP** variable.

Here, we apply the function to 5 datasets of the GEO2KEGG compendium. 

```{r deAna}
geo2kegg <- runDE(geo2kegg, de.method="limma", padj.method="flexible")
rowData(geo2kegg[[1]], use.names=TRUE)
```

Note: DE studies typically report a gene as differentially expressed
 if the corresponding DE *p*-value, corrected for multiple testing, satisfies 
the chosen significance level.
Enrichment methods that work directly on the list of DE genes are then 
substantially influenced by the multiple testing correction. 

An example is the frequently used over-representation analysis (ORA), which 
assesses the overlap between the DE genes and a gene set under study based
on the hypergeometric distribution (see the vignette of the 
`r Biocpkg("EnrichmentBrowser")` package, Appendix A, for an introduction).

ORA is inapplicable if there are few genes satisfying the significance threshold,
or if almost all genes are DE.

Using `padj.method="flexible"` accounts for these cases by applying multiple 
testing correction in dependence on the observed degree of differential expression: 

- the correction method from
    [Benjamini and Hochberg](http://sci2s.ugr.es/keel/pdf/specific/articulo/Shaffer95MHT.pdf)
    (BH) is applied, if it renders $\ge 1\%$ and $\le 25\%$ of all measured 
    genes as DE,
- the *p*-values are left unadjusted, if the BH correction results in $<1\%$ 
    DE genes, and
- the more stringent
    [Bonferroni](http://sci2s.ugr.es/keel/pdf/specific/articulo/Shaffer95MHT.pdf)
    correction is applied, if the BH correction results in $>25\%$ DE genes. 

Note that resulting $p$-values are not further used for assessing the 
statistical significance of DE genes within or between datasets. 
They are solely used to determine which genes are included in the analysis with
ORA - where the flexible correction ensures that the fraction of included genes
is roughly in the same order of magnitude across datasets. 
Alternative strategies could also be applied (such as taking a constant number of
genes for each dataset or generally excluding ORA methods from the assessment). 


# Enrichment analysis

In the following, we demonstrate how to carry out enrichment analysis in a 
benchmark setup.
Therefore, we use the collection of human KEGG gene sets as obtained 
with `getGenesets` from the `r Biocpkg("EnrichmentBrowser")`
package.

```{r getGS}
library(EnrichmentBrowser)
kegg.gs <- getGenesets(org="hsa", db="kegg")
```

At the core of applying a specific enrichment method to a single dataset is the
`runEA` function, which delegates execution of the chosen method to either
`EnrichmentBrowser::sbea` (set-based enrichment analysis) or
`EnrichmentBrowser::nbea` (network-based enrichment analysis).
In addition, it returns CPU time used and allows saving results for subsequent
assessment.

Here, we carry out ORA on the first dataset of the GEO2KEGG compendium.
```{r runEA}
kegg.ora.res <- runEA(geo2kegg[[1]], method="ora", gs=kegg.gs, perm=0)
kegg.ora.res
```

The function `runEA` can also be used to carry out several methods on multiple 
datasets.
As an example, we carry out ORA and
[CAMERA](https://doi.org/10.1093/nar/gks461)
on 5 datasets of the GEO2KEGG compendium saving the results in a temporary 
directory.

```{r eaAll}
res.dir <- tempdir()
res <- runEA(geo2kegg, methods=c("ora", "camera"), 
                gs=kegg.gs, perm=0, save2file=TRUE, out.dir=res.dir)
res$ora[1:2]
```

Note: saving the results to file is typically recommended when carrying out
several methods on multiple datasets for subsequent assessment. 
This makes results, potentially obtained from time-consuming computations, 
persistent across _R_ sessions. 
In case of unexpected errors, this also allows resumption from the point of failure.

# Benchmarking

Once methods have been applied to a chosen benchmark compendium, they can be 
subjected to a comparative assessment of runtime, statistical significance,
and phenotype relevance. 

To demonstrate how each criterion can be evaluated, we consider the example of 
the previous section where we applied ORA and CAMERA on 5 datasets of the 
GEO2KEGG compendium. 

However, note that this minimal example is used to illustrate the basic 
functionality in a time-saving manner - as generally intended in a vignette. 
To draw conclusions on the individual performance of both methods, a more 
comprehensive assessment, involving application to the full compendium, should be 
carried out.

## Runtime

Runtime, i.e. CPU time used, is an important measure of the 
applicability of a method.
For enrichment methods, runtime mainly depends on whether methods rely on 
permutation testing, and how computationally intensive recomputation of 
the respective statistic in each permutation is (see
[Figure 4](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4721010/figure/Fig4)
in
[Geistlinger et al., 2016](https://doi.org/10.1186/s12859-016-0884-1)).
  
To obtain the runtime from the application of ORA and CAMERA to 5 datasets of the
GEO2KEGG compendium, we can use the `readResults` function as we have saved 
results to the indicated result directory in the above call of `runEA`.
 
```{r readRT}
ea.rtimes <- readResults(res.dir, names(geo2kegg), 
                            methods=c("ora", "camera"), type="runtime")
ea.rtimes
```

For visualization of assessment results, the `bpPlot` function can be used to 
create customized boxplots for specific benchmark criteria.   

```{r plotRuntime, fig.width=6, fig.height=6}
bpPlot(ea.rtimes, what="runtime")
```

As both methods are simple gene set tests without permutation, they are among the
fastest in the field - with CAMERA being roughly twice as fast as ORA.

```{r runtimeORAvsCAMERA, fig.width=6, fig.height=6}
mean(ea.rtimes$ora) / mean(ea.rtimes$camera)
```

## Fraction of significant gene sets

The statistical accuracy of the significance estimation in gene set tests has 
been repeatedly debated.
For example, systematic inflation of statistical significance in ORA due to an 
unrealistic independence assumption between genes is well-documented 
([Goeman and Buehlmann, 2007](https://doi.org/10.1093/bioinformatics/btm051)).
On the other hand, the permutation procedure incorporated in many gene set tests
has been shown to be biased
([Efron and Tibshirani, 2007](https://doi.org/10.1214/07-AOAS101)),
and also inaccurate if permutation $p$-values are reported as zero 
([Phipson and Smyth, 2010](https://doi.org/10.2202/1544-6115.1585)).

These shortcomings can lead to inappropriately large fractions of significant
gene sets, and can considerably impair prioritization of gene sets in practice.
It is therefore important to evaluate resulting fractions of significant gene 
sets in comparison to other methods and with respect to the fraction of 
differentially expressed genes as a baseline.

We use the `readResults` function to obtain the saved gene set rankings of ORA
and CAMERA when applied to 5 datasets of the GEO2KEGG compendium (see above call
of `runEA`).

```{r readRankings}
ea.ranks <- readResults(res.dir, names(geo2kegg), 
                            methods=c("ora", "camera"), type="ranking")
lengths(ea.ranks)
ea.ranks$ora[1:2]
```

The `evalNrSigSets` calculates the percentage of significant gene sets given a 
significance level `alpha` and a multiple testing correction method `padj`. 
We can visualize assessment results as before using `bpPlot`, which demonstrates
 here that CAMERA produces substantially larger fractions of significant 
gene sets than ORA.

```{r plotAdjSigSets, fig.width=6, fig.height=6}
sig.sets <- evalNrSigSets(ea.ranks, alpha=0.05, padj="BH")
sig.sets
bpPlot(sig.sets, what="sig.sets")
```

## Phenotype relevance

As introduced above, Tarca et al. ([2012](https://doi.org/10.1186/1471-2105-13-136) 
and [2013](https://doi.org/10.1371/journal.pone.0079217)) also assigned a target
 pathway to each dataset of the GEO2KEGG compendium, which is considered 
highly-relevant for the respective phenotype investigated. 
However, the relation between dataset, investigated phenotype, and assigned target
pathway is not always clear-cut.
In addition, there is typically more than one pathway that is considered relevant
for the investigated phenotype.

On the other hand, evaluations of published enrichment methods often conclude on
phenotype relevance, if there is *any* association between top-ranked gene sets 
and the investigated phenotype.

A more systematic approach is used in the
[MalaCards](https://www.malacards.org) database of human diseases.
Here, relevance of [GO](http://www.geneontology.org)
and [KEGG](http://www.genome.jp/kegg) gene sets is summarized from 
(*i*) experimental evidence and (*ii*) co-citation with the respective disease 
in the literature.  

### MalaCards disease relevance rankings

The `r Biocpkg("GSEABenchmarkeR")` package provides MalaCards relevance rankings
for the diseases investigated in the datasets of the GEO2KEGG and TCGA compendia.
Here, we load the relevance rankings for KEGG gene sets and demonstrate how they
can be incorporated in the assessment of phenotype relevance.

We note that the relevance rankings contain different numbers of gene sets for 
different diseases, because only gene sets for which evidence/association with the 
respective disease has been found are listed in a ranking. 

For demonstration, we inspect the relevance rankings for Alzheimer's disease 
(ALZ) and breast cancer (BRCA) containing 57 and 142 gene sets, respectively. 

```{r malaRankings}
mala.kegg.file <- file.path(data.dir, "malacards", "KEGG.rds")
mala.kegg <- readRDS(mala.kegg.file)
sapply(mala.kegg, nrow)
mala.kegg$ALZ
mala.kegg$BRCA
```

### Mapping between dataset ID and disease code

To obtain the relevance ranking of the respective disease investigated when 
assessing results on a specific dataset, a mapping between dataset and 
investigated disease is required. 
The function `readDataId2diseaseCodeMap` reads such a mapping from a tabular 
text file and turns it into a named vector - where the elements correspond to
the disease codes and the names to the dataset IDs.

Here, we read the mapping between GSE ID and disease code for the GEO2KEGG
compendium.   

```{r data2dis}
d2d.file <- file.path(data.dir, "malacards", "GseId2Disease.txt")
d2d.map <- readDataId2diseaseCodeMap(d2d.file)
head(d2d.map)
```

### Relevance score of a gene set ranking

To evaluate the phenotype relevance of a gene set ranking obtained from the 
application of an enrichment method to an expression dataset, the function
`evalRelevance` assesses whether the ranking accumulates phenotype-relevant gene 
sets (i.e. gene sets with high relevance scores) at the top.
Therefore, the function first transforms the ranks from the enrichment analysis
to weights - where the greater the weight of a gene set, the more it is ranked 
towards the top of the GSEA ranking.
These weights are then multiplied by the corresponding relevance scores and 
summed.
 
Here, we use `evalRelevance` to assess whether ORA, when applied to the GSE1297 
dataset, recovers Alzheimer-relevant KEGG pathways.  

```{r evalRelevance}
ea.ranks$ora$GSE1297
obs.score <- evalRelevance(ea.ranks$ora$GSE1297, mala.kegg$ALZ)
obs.score
```

### Random relevance score distribution

To assess the significance of the observed relevance score of an enrichment
method applied to a specific dataset, i.e. to assess how likely it is to 
observe a relevance score equal or greater than the one obtained, the function
`compRand` repeatedly applies `evalRelevance` to randomly drawn gene set rankings. 

For demonstration, we compute relevance scores for 50 random gene set rankings
and calculate the *p*-value as for a permutation test. 
This demonstrates that the relevance score obtained from applying ORA to GSE1297
significantly exceeds random scores.

```{r compRand}
gs.names <- ea.ranks$ora$GSE1297$GENE.SET
gs.ids <- substring(gs.names, 1, 8)
rand.scores <- compRand(mala.kegg$ALZ, gs.ids, perm=50)
summary(rand.scores)
(sum(rand.scores >= obs.score) + 1) / 51
```

### Theoretical optimum

The observed relevance score can be used to compare phenotype relevance of 
two or more methods when applied to one particular dataset.
However, as the number of gene sets in the relevance rankings differs between 
phenotypes (see above Section *5.3.1 MalaCards disease relevance rankings*), 
comparison between datasets is not straightforward as resulting relevance scores
scale differently. 

Therefore, the function `compOpt` applies `evalRelevance` to the theoretically
optimal case in which the enrichment analysis ranking is identical to the 
relevance score ranking. The ratio between observed and optimal score can then
be used to compare observed scores between datasets.

Here, we compute the optimal score for the Alzheimer relevance ranking, which
indicates that the score observed for ORA, when applied to GSE1297, is about 68%
of the optimal score.

```{r compOpt}
opt.score <- compOpt(mala.kegg$ALZ, gs.ids)
opt.score
round(obs.score / opt.score * 100, digits=2)
```

### Cross-dataset relevance score distribution

Evaluation of phenotype relevance with `evalRelevance` can also be done for 
several methods applied across multiple datasets.
This allows to assess whether certain enrichment methods tend to produce 
rankings of higher phenotype relevance than other methods when applied to a
compendium of datasets.
As explained in the previous section, observed relevance scores are always 
expressed in relation to the respective optimal score.

For demonstration, we use `evalRelevance` to evaluate phenotype relevance of the
gene set rankings produced by ORA and CAMERA when applied to 5 datasets of the
GEO2KEGG compendium.
We can visualize assessment results as before using `bpPlot`, which demonstrates
here that ORA tends to recover more phenotype-relevant gene sets than CAMERA.  

```{r evalAll, fig.width=6, fig.height=6}
all.kegg.res <- evalRelevance(ea.ranks, mala.kegg, d2d.map[names(geo2kegg)])
bpPlot(all.kegg.res, what="rel.sets")
```

# Advanced 

## Caching

Preparing an expression data compendium for benchmarking of enrichment methods
can be time-consuming.
In case of the GEO2KEGG compendium, it requires to summarize probe level 
expression on gene level and to subsequently carry out differential expression 
analysis for each dataset.

To flexibly save and restore an already processed expression data compendium, 
we can use the `cacheResource` function which builds on functionality of the
`r Biocpkg("BiocFileCache")` package.

```{r cacheRes}
cacheResource(geo2kegg, rname="geo2kegg")
```

This adds the selected 5 datasets of the GEO2KEGG compendium (as processed 
throughout this vignette) to the cache, and allows to restore it at a later time
via

```{r getRes}
geo2kegg <- loadEData("geo2kegg", cache=TRUE)
names(geo2kegg)
```

Note: to obtain the original unprocessed version of the compendium, set the 
`cache` argument of the `loadEData` function to `FALSE`.

To clear the cache (use with care):

```{r clearCache, eval=FALSE}
cache.dir <- rappdirs::user_cache_dir("GSEABenchmarkeR")
bfc <- BiocFileCache::BiocFileCache(cache.dir)
BiocFileCache::removebfc(bfc)
```

## Parallel computation

Leveraging functionality from `r Biocpkg("BiocParallel")`, parallel computation
of the functions `maPreproc`, `runDE`, and
especially `runEA`, when applied to multiple datasets is straightforward.
Internally, these functions call `BiocParallel::bplapply`, which triggers parallel
computation as configured in the first element of `BiocParallel::registered()`.
As a result, parallel computation is implicitly incorporated in the above calls
of these functions when carried out on a multi-core machine.
See the vignette of the `r Biocpkg("BiocParallel")` package for an introduction.

Inspecting 
```{r bpRegister}
BiocParallel::registered()
```
shows that the execution uses a `MulticoreParam` per default (on Windows:
a `SnowParam`), where the `bpnworkers` attribute indicates the number of cores 
involved in the computation.

To change the execution mode of functions provided in the
`r Biocpkg("GSEABenchmarkeR")` package, accordingly configured computation 
parameters of class `BiocParallelParam` can either directly be registered via
`BiocParallel::register`, or supplied with the `parallel` argument of the
respective function.

For demonstration, we configure here a `BiocParallelParam` to display a progress
bar

```{r bpParam}
bp.par <- BiocParallel::registered()[[1]]
BiocParallel::bpprogressbar(bp.par) <- TRUE
```

and supply `runDE` with the updated computation parameter.

```{r runDEBP}
geo2kegg <- runDE(geo2kegg, parallel=bp.par)
```

Users that would like to use distributed computation, on e.g. an institutional 
computer cluster, should consult the vignette of the `r Biocpkg("BiocParallel")`
package to similarly configure a `BiocParallelParam` for that purpose.
